---
title: "Maps with tmap"
author: "Rexford Osei Owusu"
date: '2023-06-20'
output: html_document
---

## Loading the tmap package
```{r}
library(tmap)
library(sf)
```

Loading the NY8 data, and subset Syracuse
```{r}
NY8 <- st_read("NY_data/NY8_utm18.shp")
```

```{r}
Syracuse <- NY8[NY8$AREANAME == "Syracuse city", ]
plot(Syracuse[, "POP8"])
```

Using tmap to make simple map showing the polygon outlines using tm_borders()
```{r}
tm_shape(Syracuse) + tm_borders()
```

Using tm_fill() will then fill these using one of the variables in the Syracuse data set (POP8)
```{r}
tm_shape(Syracuse) + tm_borders() + tm_fill("POP8")
```

Changing color setting by the palette argument in tm_fill()
```{r}
tm_shape(Syracuse) + tm_borders() + 
  tm_fill("POP8", palette = "Blues", style = "quantile")
```

Adding elements such as longitude/latitude graticule with tm_graticules(), a north arrow and a line of text with the date the map was made
```{r}
tm_shape(Syracuse) + tm_graticules(col = "lightgray") + tm_borders() + 
  tm_fill("POP8", palette = "Greens", style = "quantile") + 
  tm_compass(position = c("left", "bottom")) + 
  tm_credits("2023-06-21", position = c("right", "top"))
```

## Interactive maps
Resetting tmap mode from static to interactive (view) and making a simple plot of the Syracuse data
```{r}
tmap_mode("view")
names(Syracuse)
tm_shape(Syracuse) + tm_borders() + tm_fill("Cases", palette = "Greens")
tmap_mode("plot")
```

Reading the Western North American site data
```{r}
wna_climate <- read.csv("WNAclimate.csv")
str(wna_climate)
```

Converting the data to a sf object (a points data frame)
```{r}
wna_climate <- st_as_sf(wna_climate, 
                        coords = c("LONDD", "LATDD"),
                        crs = 4326)
```

Plotting individual symbols on a color scale using tm_symbols
```{r}
tm_shape(wna_climate) + tm_symbols(col="Jan_Tmp")
```

Using the tm_fill() argument for the color palette. The minus sign before the palette name reverses the order of the colors. The alpha level is added to make the symbols transparent.
```{r}
tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.5, palette = "-RdBu") 
```

Loading the Natural Earth shapefile of country boundaries
```{r}
countries <- st_read("ne_50m_admin_0_countries/ne_50m_admin_0_countries.shp")
```

Adding the shapefile as a different spatial object, we use tm_shape() a second time to reference this, then use tm_borders() to add the lines
```{r}
tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.75, palette = "-RdBu") + 
  tm_shape(countries) + tm_borders(col="gray")
```

Using tm_style() to alter the appearance of the map
```{r}
tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.75, palette = "-RdBu", size = 0.5) + 
  tm_shape(countries) + tm_borders(col="gray") + tm_style("cobalt") 
```

Natural style
```{r}
tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.75, palette = "-RdBu", size = 0.5) + 
  tm_shape(countries) + tm_borders(col="red") + tm_style("natural") 
```

Classic style
```{r}
tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.75, palette = "-RdBu", size = 0.5) + 
  tm_shape(countries) + tm_borders(col="gray") + tm_style("classic") 
```

Albatross style
```{r}
tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.75, palette = "-RdBu", size = 0.5) + 
  tm_shape(countries) + tm_borders(col="yellow") + tm_style("albatross") 
```

## Making two maps (January and July temperature) into R objects and saving them as tm1 and tm2
```{r}
tm1 <- tm_shape(wna_climate) + 
  tm_symbols(col="Jan_Tmp", alpha = 0.75, palette = "-RdBu", size = 0.5) + 
  tm_shape(countries) + tm_borders(col="blue") + tm_style("classic") 

tm2 <- tm_shape(wna_climate) + 
  tm_symbols(col="Jul_Tmp", alpha = 0.75, palette = "-RdBu", size = 0.5) + 
  tm_shape(countries) + tm_borders(col="orange") + tm_style("watercolor") 
```

Two maps can be made in a single figure using tm_arrange()
```{r}
tmap_arrange(tm1, tm2)
```

## Making figures using the global air temperature dataset
Reading the data using the raster package
```{r}
library(raster)
library(ncdf4)
r <- raster("air.mon.ltm.nc", varname="air")
r <- rotate(r)
names(r)
```

Changing variable names in our raster layer to something simpler
```{r}
names(r) <- "jan_tmp"
```

Plotting using tm_raster(). This takes color palettes and borders
```{r}
tm_shape(r) + 
  tm_raster("jan_tmp", style="fisher", palette="-RdBu") +
  tm_shape(countries) + tm_borders() 
```

Setting the projection of the raster r using the crs() function
```{r}
crs(r) <- 4326
```

Remaking the figure with the projected raster r
```{r}
tm_shape(r) + 
  tm_raster("jan_tmp", palette="RdBu") +
  tm_shape(countries) + tm_borders() 
```

Reversing the palette by prepending a - before the palette name. Adding n=9 to the function to increase the number of intervals to see spatial variation and adding background to make legend visible
```{r}
tm_shape(r) + 
  tm_raster("jan_tmp", style="fisher", palette="-RdBu", n = 9, title = "January temperature") +
  tm_shape(countries) + tm_borders() + 
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.6)
```

Alternatively,by moving the color legend outside of the plotting area makes it more visible. We’ll also increase the number of color classes to 9, and add a histogram showing the frequency of different values to see the spatial variations.
```{r}
tm_shape(r) + 
  tm_raster("jan_tmp", style="fisher", palette="-RdBu", legend.hist = TRUE, 
            n = 9, title = "January temperature") +
  tm_shape(countries) + tm_borders() + tm_layout(legend.outside = TRUE, 
                                                 legend.outside.position = "left")
```

